---
title: "Example for msma"
output:
  html_document:
    toc: true
    toc_float: true
  pdf_document:
    toc: true
    highlight: zenburn
  rmarkdown::html_vignette:
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{msma}
  %\VignetteEncoding{UTF-8}
---

<!--
%\VignetteEngine{knitr::knitr}
%\\VignetteIndexEntry{msma}
#source("G:/ADNI/Multimodal/program/package/Ver1/src.r"); library(mvtnorm)
source("C:/Users/kawaguchi/ADNI/Multimodal/program/package/Ver1/src.r"); library(mvtnorm)
-->

`r Sys.Date()` @Atsushi Kawaguchi

# Preparation

Install package (as necessary)

```{r eval=FALSE}
if(!require("msma")) install.packages("msma")
```

Load package

```{r}
library(msma)
```

# Getting started
Simulated multiblock data (list data) by using the function `simdata`.
Sample size is 50. Response has three blocks and predictor has one block.
```{r}
dataset0 = simdata(n = 50, rho = 0.8, Yps = c(3, 4, 5), Xps = 3, seed=1)
X0 = dataset0$X; Y0 = dataset0$Y 
```

The argument comp can specify the number of components.

- One Component
```{r}
fit01 = msma(X0, Y0, comp=1, lambdaX=0.05, lambdaY=1:3)
fit01
plot(fit01)
```

- Two Component
```{r}
fit02 = msma(X0, Y0, comp=2, lambdaX=0.03, lambdaY=0.01*(1:3))
fit02
```

# Single Block

```{r}
dataset1 = simdata(n = 50, rho = 0.8, Yps = 5, Xps = 5, seed=1)
X1 = dataset1$X[[1]]; Y1 = dataset1$Y 
```

## Principal Component Analysis (PCA)
If input is a matrix, a principal component analysis is implemented.
```{r}
(fit111 = msma(X1, comp=5))
```

The weight (loading) vectors
```{r}
fit111$wbX
```

The bar plots of weight vectors 
```{r}
par(mfrow=c(1,2))
plot(fit111, axes = 1, plottype="bar")
plot(fit111, axes = 2, plottype="bar")
```

The score vectors for six subjects.
```{r}
lapply(fit111$sbX, head)
```

The scatter plots for the score vectors.
```{r}
plot(fit111, v="score", axes = 1:2, plottype="scatter")
plot(fit111, v="score", axes = 2:3, plottype="scatter")
```

```{r}
par(mfrow=c(1,1))
plot(fit111, v="cpev")
```
### Other functions in R for PCA
There is the R function prcomp to implement PCA.
```{r}
(fit1112 = prcomp(X1, scale=TRUE))
summary(fit1112)
biplot(fit1112)
```

The `ggfortify` package is also available for the PCA plot.

### Sparse PCA
If `lambdaX` (>0) is specified, a sparse principal component analysis is implemented.
```{r}
(fit112 = msma(X1, comp=5, lambdaX=0.1))
par(mfrow=c(1,2))
plot(fit112, axes = 1, plottype="bar")
plot(fit112, axes = 2, plottype="bar")
```

### Supervised Sparse PCA
If the outcome Z is specified, a supervised sparse principal component analysis is implemented.
```{r}
set.seed(1); Z = rbinom(50, 1, 0.5)

(fit113 = msma(X1, Z=Z, comp=5, lambdaX=0.02))
par(mfrow=c(1,2))
plot(fit113, axes = 1, plottype="bar")
plot(fit113, axes = 2, plottype="bar")
```

## Partial Least Squres (PLS)
If the another input is specified, a partial least squres is implemented.
```{r}
(fit121 = msma(X1, Y1, comp=2))
par(mfrow=c(1,2))
plot(fit121, axes = 1, XY="XY")
plot(fit121, axes = 2, XY="XY")
```

### Sparse PLS
If `lambdaX` and `lambdaY` are specified, a sparse PLS is implemented.
```{r}
(fit122 = msma(X1, Y1, comp=2, lambdaX=0.5, lambdaY=0.5))
par(mfrow=c(1,2))
plot(fit122, axes = 1, XY="XY")
plot(fit122, axes = 2, XY="XY")
```

### Supervised Sparse PLS
If the outcome Z is specified, a supervised sparse PLS is implemented.
```{r}
(fit123 = msma(X1, Y1, Z, comp=2, lambdaX=0.5, lambdaY=0.5))
par(mfrow=c(1,2))
plot(fit123, axes = 1, XY="XY")
plot(fit123, axes = 2, XY="XY")
```


# Multi Block
Multiblock data is a list of data matrix. 
```{r}
dataset2 = simdata(n = 50, rho = 0.8, Yps = c(2, 3), Xps = c(3, 4), seed=1)
X2 = dataset2$X; Y2 = dataset2$Y 
```

## PCA

The input class is list.
```{r}
class(X2)
```

The list length is 2 for 2 blocks.
```{r}
length(X2)
```

list of data matrix structure.
```{r}
lapply(X2, dim)
```

```{r}
(fit211 = msma(X2, comp=1))
```

The bar plots for the block and super weights (loadings). 
```{r}
par(mfrow=c(1,2))
plot(fit211, axes = 1, plottype="bar", block="block")
plot(fit211, axes = 1, plottype="bar", block="super")
```

### Sparse PCA
If `lambdaX` with the length of 2 (same as the length of blocks) are specified, a multiblock sparse PCA is implemented.
```{r}
(fit212 = msma(X2, comp=1, lambdaX=c(0.5, 0.5)))
```


```{r}
par(mfrow=c(1,2))
plot(fit212, axes = 1, plottype="bar", block="block")
plot(fit212, axes = 1, plottype="bar", block="super")
```

### Supervised Sparse PCA
```{r}
(fit213 = msma(X2, Z=Z, comp=1, lambdaX=c(0.5, 0.5)))
par(mfrow=c(1,2))
plot(fit213, axes = 1, plottype="bar", block="block")
plot(fit213, axes = 1, plottype="bar", block="super")
```

## PLS
If the another input is specified, the partial least squared is implemented.
```{r}
(fit221 = msma(X2, Y2, comp=1))
par(mfrow=c(1,2))
plot(fit221, axes = 1, plottype="bar", block="block", XY="X")
plot(fit221, axes = 1, plottype="bar", block="super", XY="X")
plot(fit221, axes = 1, plottype="bar", block="block", XY="Y")
plot(fit221, axes = 1, plottype="bar", block="super", XY="Y")
```

### Sparse PLS
The regularized parameters
```{r}
(fit222 = msma(X2, Y2, comp=1, lambdaX=c(0.5, 0.5), lambdaY=c(0.5, 0.5)))
par(mfrow=c(1,2))
plot(fit222, axes = 1, plottype="bar", block="block", XY="X")
plot(fit222, axes = 1, plottype="bar", block="super", XY="X")
plot(fit222, axes = 1, plottype="bar", block="block", XY="Y")
plot(fit222, axes = 1, plottype="bar", block="super", XY="Y")
```

### Supervised Sparse PLS
```{r}
(fit223 = msma(X2, Y2, Z, comp=1, lambdaX=c(0.5, 0.5), lambdaY=c(0.5, 0.5)))
par(mfrow=c(1,2))
plot(fit223, axes = 1, plottype="bar", block="block", XY="X")
plot(fit223, axes = 1, plottype="bar", block="super", XY="X")
plot(fit223, axes = 1, plottype="bar", block="block", XY="Y")
plot(fit223, axes = 1, plottype="bar", block="super", XY="Y")
```

# Parameter Selection

## PCA
```{r}
criteria = c("BIC", "CV")
search.methods = c("regparaonly", "regpara1st", "ncomp1st", "simultaneous")
c1=criteria[1]; sm1=search.methods[1]
(opt1 = optparasearch(X1, search.method = sm1, criterion=c1))
(fit31 = msma(X1, comp=opt1$optncomp, lambdaX=opt1$optlambdaX))
```

## PLS
```{r}
(opt2 = optparasearch(X2, Y2, search.method = sm1, criterion=c1))
(fit32 = msma(X2, Y2, comp=opt2$optncomp, lambdaX=opt2$optlambdaX, lambdaY=opt2$optlambdaY))
```
