---
title: "Example for msma"
output:
  html_document:
    toc: true
    toc_float: true
    fig_width: 6
    fig_height: 5
  rmarkdown::html_vignette:
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{msma}
  %\VignetteEncoding{UTF-8}
---

<!--
%\VignetteEngine{knitr::knitr}
%\\VignetteIndexEntry{msma}
-->

`r Sys.Date()` @Atsushi Kawaguchi

The `msma` package provides functions for a matrix decomposition method incorporating sparse and supervised modeling for a multiblock multivariable data analysis. 

# Preparation

Install package (as necessary)

```{r eval=FALSE}
if(!require("msma")) install.packages("msma")
```

Load package

```{r}
library(msma)
```

```{r echo=FALSE, error=FALSE}
#predir = unlist(strsplit(getwd(), "ADNI"))[1]
#source(paste(predir, "ADNI/Multimodal/program/package/Ver2/src.r", sep="")); library(mvtnorm)
```

<!--
source("G:/ADNI/Multimodal/program/package/Ver2/src.r")
source("C:/Users/kawaguchi/ADNI/Multimodal/program/package/Ver2/src.r")
-->

# Getting started
Simulated multiblock data (list data) by using the function `simdata`.
Sample size is 50. The correlation coeficient is 0.8. 
The numbers of columns for response and predictor can be specified by the argument `Yps` and `Xps`, respectively.
The length of vecor represents the number of blocks.
That is, response has three blocks with the numbers of columns being 3, 4, and 5 and predictor has one block with the number of columns being 3.
```{r}
dataset0 = simdata(n = 50, rho = 0.8, Yps = c(3, 4, 5), Xps = 3, seed=1)
X0 = dataset0$X; Y0 = dataset0$Y 
```

The argument `comp` can specify the number of components.
The arguments `lambdaX` and `lambdaY` can specify the regularization parameters for X and Y, respectively.

- One Component
```{r}
fit01 = msma(X0, Y0, comp=1, lambdaX=0.05, lambdaY=1:3)
fit01
```

The `plot` function is available. 
In default setting, the block weights are displayed as a barplot.
```{r}
plot(fit01)
```

- Two Component
```{r}
fit02 = msma(X0, Y0, comp=2, lambdaX=0.03, lambdaY=0.01*(1:3))
fit02
```

################################################################################################
# Single Block
################################################################################################

Two matrics are prepared by specifying arguments `Yps` and `Xps`.
```{r}
dataset1 = simdata(n = 50, rho = 0.8, Yps = 5, Xps = 5, seed=1)
X1 = dataset1$X[[1]]; Y1 = dataset1$Y 
```

## Principal Component Analysis (PCA)
If input is a matrix, a principal component analysis is implemented.
```{r}
(fit111 = msma(X1, comp=5))
```

The weight (loading) vectors can be obtained as follows.
```{r}
fit111$wbX
```

The bar plots of weight vectors are provided by the function `plot`.
The component number is specified by the argument `axes`.
The plot type is selected by the argument `plottype`.
```{r}
par(mfrow=c(1,2))
plot(fit111, axes = 1, plottype="bar")
plot(fit111, axes = 2, plottype="bar")
```

The score vectors for first six subjects.
```{r}
lapply(fit111$sbX, head)
```

The scatter plots for the score vectors specified by the argument `v`.
The argument `axes` is specified by the two length vector represents which components are displayed.
```{r}
plot(fit111, v="score", axes = 1:2, plottype="scatter")
plot(fit111, v="score", axes = 2:3, plottype="scatter")
```

When the argument `v` was specified as "cpev", the cummulative eigenvalues are plotted. 
```{r}
par(mfrow=c(1,1))
plot(fit111, v="cpev")
```

### Other functions in R for PCA
There is the R function prcomp to implement PCA.
```{r}
(fit1112 = prcomp(X1, scale=TRUE))
summary(fit1112)
biplot(fit1112)
```

The `ggfortify` package is also available for the PCA plot.


### Sparse PCA
If `lambdaX` (>0) is specified, a sparse principal component analysis is implemented.
```{r}
(fit112 = msma(X1, comp=5, lambdaX=0.1))
par(mfrow=c(1,2))
plot(fit112, axes = 1, plottype="bar")
plot(fit112, axes = 2, plottype="bar")
```

### Supervised Sparse PCA
The outcome Z is generated.
```{r}
set.seed(1); Z = rbinom(50, 1, 0.5)
```

If the outcome Z is specified, a supervised sparse principal component analysis is implemented.
```{r}
(fit113 = msma(X1, Z=Z, comp=5, lambdaX=0.02))
```

```{r}
par(mfrow=c(1,2))
plot(fit113, axes = 1, plottype="bar")
plot(fit113, axes = 2, plottype="bar")
```

## Partial Least Squres (PLS)
If the another input Y1 is specified, a partial least squres is implemented.
```{r}
(fit121 = msma(X1, Y1, comp=2))
```

The component number is specified by the argument `axes`.
When the argument `XY` was specified as "XY", the scatter plots for Y score against X score are plotted. 
```{r}
par(mfrow=c(1,2))
plot(fit121, axes = 1, XY="XY")
plot(fit121, axes = 2, XY="XY")
```


### Sparse PLS
If `lambdaX` and `lambdaY` are specified, a sparse PLS is implemented.
```{r}
(fit122 = msma(X1, Y1, comp=2, lambdaX=0.5, lambdaY=0.5))
par(mfrow=c(1,2))
plot(fit122, axes = 1, XY="XY")
plot(fit122, axes = 2, XY="XY")
```

### Supervised Sparse PLS
If the outcome Z is specified, a supervised sparse PLS is implemented.
```{r}
(fit123 = msma(X1, Y1, Z, comp=2, lambdaX=0.5, lambdaY=0.5))
par(mfrow=c(1,2))
plot(fit123, axes = 1, XY="XY")
plot(fit123, axes = 2, XY="XY")
```


################################################################################################
# Multi Block
################################################################################################
Multiblock data is a list of data matrix. 
```{r}
dataset2 = simdata(n = 50, rho = 0.8, Yps = c(2, 3), Xps = c(3, 4), seed=1)
X2 = dataset2$X; Y2 = dataset2$Y 
```

## PCA

The input class is list.
```{r}
class(X2)
```

The list length is 2 for 2 blocks.
```{r}
length(X2)
```

list of data matrix structure.
```{r}
lapply(X2, dim)
```

The function `msma` is applied to this list X2 as follows.
```{r}
(fit211 = msma(X2, comp=1))
```

The bar plots for the block and super weights (loadings) specified the argument `block`. 
```{r}
par(mfrow=c(1,2))
plot(fit211, axes = 1, plottype="bar", block="block")
plot(fit211, axes = 1, plottype="bar", block="super")
```

### Sparse PCA
If `lambdaX` with the length of 2 (same as the length of blocks) are specified, a multiblock sparse PCA is implemented.
```{r}
(fit212 = msma(X2, comp=1, lambdaX=c(0.5, 0.5)))
```

The bar plots for the block and super weights (loadings). 
```{r}
par(mfrow=c(1,2))
plot(fit212, axes = 1, plottype="bar", block="block")
plot(fit212, axes = 1, plottype="bar", block="super")
```

### Supervised Sparse PCA
If the outcome Z is specified, a supervised analysis is implemented.
```{r}
(fit213 = msma(X2, Z=Z, comp=1, lambdaX=c(0.5, 0.5)))
par(mfrow=c(1,2))
plot(fit213, axes = 1, plottype="bar", block="block")
plot(fit213, axes = 1, plottype="bar", block="super")
```

## PLS
If the another input (list) Y2 is specified, the partial least squared is implemented.
```{r}
(fit221 = msma(X2, Y2, comp=1))
```

```{r}
par(mfrow=c(1,2))
plot(fit221, axes = 1, plottype="bar", block="block", XY="X")
plot(fit221, axes = 1, plottype="bar", block="super", XY="X")
```

```{r}
par(mfrow=c(1,2))
plot(fit221, axes = 1, plottype="bar", block="block", XY="Y")
plot(fit221, axes = 1, plottype="bar", block="super", XY="Y")
```

### Sparse PLS
The regularized parameters `lambdaX` and `lambdaY` are specified vectors with same length with the length of lists X2 and Y2, respectively.
```{r}
(fit222 = msma(X2, Y2, comp=1, lambdaX=c(0.5, 0.5), lambdaY=c(0.5, 0.5)))
```

```{r}
par(mfrow=c(1,2))
plot(fit222, axes = 1, plottype="bar", block="block", XY="X")
plot(fit222, axes = 1, plottype="bar", block="super", XY="X")
```


```{r}
par(mfrow=c(1,2))
plot(fit222, axes = 1, plottype="bar", block="block", XY="Y")
plot(fit222, axes = 1, plottype="bar", block="super", XY="Y")
```

### Supervised Sparse PLS
```{r}
(fit223 = msma(X2, Y2, Z, comp=1, lambdaX=c(0.5, 0.5), lambdaY=c(0.5, 0.5)))
```

```{r}
par(mfrow=c(1,2))
plot(fit223, axes = 1, plottype="bar", block="block", XY="X")
plot(fit223, axes = 1, plottype="bar", block="super", XY="X")
```

```{r}
par(mfrow=c(1,2))
plot(fit223, axes = 1, plottype="bar", block="block", XY="Y")
plot(fit223, axes = 1, plottype="bar", block="super", XY="Y")
```

################################################################################################
# Parameter Selection
################################################################################################

The number of components and regularized parameters can be selected by the function `optparasearch`.
The following options are available.

```{r}
criteria = c("BIC", "CV")
search.methods = c("regparaonly", "regpara1st", "ncomp1st", "simultaneous")
```

## Single Block

### PCA

-  The `regparaonly` method searches for the regularized parameters with a fixed number of components.

```{r}
(opt11 = optparasearch(X1, search.method = "regparaonly", criterion="BIC"))
(fit311 = msma(X1, comp=opt11$optncomp, lambdaX=opt11$optlambdaX))
```

-  The `regpara1st` identifies the regularized parameters by fixing the number of components, then searching for the number of components with the selected regularized parameters. 

```{r}
(opt12 = optparasearch(X1, search.method = "regpara1st", criterion="BIC"))
(fit312 = msma(X1, comp=opt12$optncomp, lambdaX=opt12$optlambdaX))
```

-  The `ncomp1st` method identifies the number of components with a regularized parameter of 0, then searches for the regularized parameters with the selected number of components. 

```{r}
(opt13 = optparasearch(X1, search.method = "ncomp1st", criterion="BIC"))
(fit313 = msma(X1, comp=opt13$optncomp, lambdaX=opt13$optlambdaX))
```

-  The `simultaneous` method identifies the number of components by searching the regularized parameters in each component. 

```{r}
(opt14 = optparasearch(X1, search.method = "simultaneous", criterion="BIC"))
(fit314 = msma(X1, comp=opt14$optncomp, lambdaX=opt14$optlambdaX))
```


The argument maxpct4ncomp=0.5 means that 0.5$\lambda$ is used as the regularized
parameter when the number of components is searched and where $\lambda$ is the maximum of
the regularized parameters among the possible candidates.

```{r}
(opt132 = optparasearch(X1, search.method = "ncomp1st", criterion="BIC", maxpct4ncomp=0.5))
(fit3132 = msma(X1, comp=opt132$optncomp, lambdaX=opt132$optlambdaX))
```


### PLS
For PLS, two parameters $\lambda_X$ and $\lambda_Y$ are used in arguments `lambdaX` and `lambdaY` to control sparseness for data X and Y, respectively.


```{r}
(opt21 = optparasearch(X2, Y2, search.method = "regparaonly", criterion="BIC"))
(fit321 = msma(X2, Y2, comp=opt21$optncomp, lambdaX=opt21$optlambdaX, lambdaY=opt21$optlambdaY))
```

## Multi Block

The multi block structure has
```{r}
dataset3 = simdata(n = 50, rho = 0.8, Yps = rep(4, 5), Xps = rep(4, 5), seed=1)
X3 = dataset3$X; Y3 = dataset3$Y 
```

### PCA

```{r}
(opt31 = optparasearch(X3, search.method = "regparaonly", criterion="BIC"))
(fit331 = msma(X3, comp=opt31$optncomp, lambdaX=opt31$optlambdaX, lambdaXsup=opt31$optlambdaXsup))
```

```{r}
(opt32 = optparasearch(X3, search.method = "regparaonly", criterion="BIC", whichselect="X"))
(fit332 = msma(X3, comp=opt32$optncomp, lambdaX=opt32$optlambdaX, lambdaXsup=opt32$optlambdaXsup))
```

```{r}
(opt33 = optparasearch(X3, search.method = "regparaonly", criterion="BIC", whichselect="Xsup"))
(fit333 = msma(X3, comp=opt33$optncomp, lambdaX=opt33$optlambdaX, lambdaXsup=opt33$optlambdaXsup))
```

### PLS

This is computationally expensive and takes much longer to execute due to the large number of blocks.
```{r, eval=FALSE}
(opt41 = optparasearch(X3, Y3, search.method = "regparaonly", criterion="BIC"))
(fit341 = msma(X3, Y3, comp=opt41$optncomp, lambdaX=opt41$optlambdaX, lambdaY=opt41$optlambdaY, lambdaXsup=opt41$optlambdaXsup, lambdaYsup=opt41$optlambdaYsup))
```

In this example, it works by narrowing down the parameters as follows.
```{r}
(opt42 = optparasearch(X3, Y3, search.method = "regparaonly", criterion="BIC", whichselect=c("Xsup","Ysup")))
(fit342 = msma(X3, Y3, comp=opt42$optncomp, lambdaX=opt42$optlambdaX, lambdaY=opt42$optlambdaY, lambdaXsup=opt42$optlambdaXsup, lambdaYsup=opt42$optlambdaYsup))
```

Another example dataset is generated.
```{r}
dataset4 = simdata(n = 50, rho = 0.8, Yps = rep(4, 2), Xps = rep(4, 3), seed=1)
X4 = dataset4$X; Y4 = dataset4$Y 
```

With this number of blocks, the calculation can be performed in a relatively short time.
```{r}
(opt43 = optparasearch(X4, Y4, search.method = "regparaonly", criterion="BIC"))
(fit343 = msma(X4, Y4, comp=opt43$optncomp, lambdaX=opt43$optlambdaX, lambdaY=opt43$optlambdaY, lambdaXsup=opt43$optlambdaXsup, lambdaYsup=opt43$optlambdaYsup))
```


